Expert-driven approximation of MPE

### 2019-08-06 methodology/novel-contributions.tex
Literature Review

\subsection{Theory} \label{subsec:theory}
\todo{la discussione con l'esperto ci da' belief revision dei dati}
\todo{counterfactual explanations}
\todo{magari togliere theory e mettere tutto assieme}

\subsubsection{Selection based on entropy}
\todo{utilizzo entropia}

\subsection{Algorithms}
An important part of my work was developing the algorithms needed to adapt the ideas presented in the paper \enquote{Explaining the Most Probable Explanation} by \cite{Butz2018} and \enquote{A Progressive Explanation of Inference in \enquote{Hybrid} Bayesian} by \cite{Kyrimi2016}.
From the former, the construction of the probability tree through a constructive dialogue with the domain expert, the building of counterfactual explanation branches, the automatic generation of the most probable probability tree from initial evidence.
From the latter, the generation of an \enquote{Inverse explanation}.
Finally, a simple procedure to output a natural language explanation was developed.

\subsubsection{\enquote{Pseudo-MPE}} \label{subsubsec:pseudo-mpe}
\todo{dove critico il fatto che non calcolano veramente l'MPE come dicevano? qui o in cap.4 on cap.2?}
The so-called \enquote{Pseudo-MPE} algorithm is inherently wrapped up with the concept of \textit{dialogue} and is central to the explanatory powers of the system being developed in this thesis.
The algorithm was developed as a way of implementing the \enquote{MPE branch} of the \enquote{Argumentative Probability Tree} hypothesised by \cite{Butz2018}.
It was termed \enquote{Pseudo-MPE} because there are no guarantees of it returning the MPE solution (see Subsec. \ref{subsec:bnupdating}), as noted by \cite{koller2007introduction} in their definition of the MAP problem.

@@

\todo{ancora da implementare true MPE}
\subsubsection{\enquote{Pseudo-MPE} from Random Evidence} \label{subsubsec: pseudo-mpe-random}
In order to compare the Pseudo-MPE output with the true MPE solution I implemented a simple algorithm that, starting from a random initial set of evidence, generates the relative pseudo-MPE and MPE solutions.
The random initial evidence set is constructed by randomly choosing a number in the interval $k = [ 1, |V| ]$, with $V$ the set of vertices in the BN, and then randomly selecting $k$ of the random variables in $V$ to yield the set of variables $E$.
The value of each variable is randomly chosen among the set of its values; as all variables are categorical, this can easily be done.

@@

The pseudocode is shown in Alg. \ref{alg:pseudo-mpe-random-evidence} while an example output can be seen in Fig. \ref{fig:pseudo-mpe-random}.

\begin{algorithm}[htp!]
	\caption{Pseudo-MPE from random evidence algorithm}
	\label{alg:pseudo-mpe-random-evidence}
	\begin{algorithmic}[1]
		\State $evidence = \emptyset$
		\State generate random number $k \in [ 1, |V| ]$
		\State $S = $ choose $k$ variables from $V$
		\For{$s \in S$}
			\State choose random $v$ in the possible values of $e$ 
			\State $evidence = evidence \cup (s,v)$
		\EndFor
		\State $MPE\_polytree = $ MPE Polytree rooted in $evidence$
		\State $last\_node = evidence$
		\State $alternative\_evidence = evidence$ 
		\While{True} 
			\State $mpe\_states$ = \texttt{next\_most\_probable\_states($alternative\_evidence$)}
			\If{$mpe\_states$ is empty}
				\State return
			\Else
				\State $next\_state$ = head of $mpe\_states$ 
				\State create $next\_state$ node and make it son of $last\_node$
				\State update $alt\_node$ node to be $next\_state$ node
				\State $alternative\_evidence = alternative\_evidence \cup next\_state$
			\EndIf
		\EndWhile
	\end{algorithmic}
\end{algorithm}

\begin{figure}[htbp]
\centerline{\includegraphics[width=\columnwidth]{methodology/images/pseudo-mpe-random-example}}
\caption{Example output of the Pseudo-MPE from random evidence algorithm.}
\label{fig:pseudo-mpe-random}
\end{figure}

@@

\subsubsection{Natural Language Explanation}
\todo{da fare e magari pensare anche a visual explanations?}
The probability of each proposed tuple is quantified in natural language based on the probability of the most probable value within the variable.
These are shown in Tab. \ref{tab:naturallanguageprobabilities}.

\begin{table*}[htbp]
\caption{Probability quantifiers in natural language}
\begin{tabularx}{\textwidth}{@{} X X @{}}
\toprule 
Probability range & Natural language quantifier \\
\midrule 
(0, 0.2) &  "highly unlikely" \\
(0.2, 0.3) & "very unlikely" \\
(0.3, 0.4) & "unlikely" \\
(0.4, 0.5) & "not plausibly" \\
(0.5, 0.6) & "plausibly" \\
(0.6, 0.7) & "possibly" \\
(0.7, 0.8) & "likely" \\
(0.8, 0.9) & "very likely" \\
(0.9, 1) &  "highly likely" \\
(1) &  "certain" \\
\bottomrule
\end{tabularx}
\label{tab:naturallanguageprobabilities}
\end{table*}

\subsubsection{Pairwise Correlations}
An interesting addition, in terms of both explainability and theory, is an algorithm to measure and graphically display the interrelatedness between pairs of variables.
This is achieved by calculating the \textit{conditional mutual information} \todo{AA: e' giusto così?} between each pair of parent $\rightarrow$ child variables.
The definition of mutual information presented in Subsec. \ref{subsec:mutualinformation} is extended to account for the current state of the model i.e. the set of observed variables.
What we are calculating is:
\begin{equation}
	I(X,Y|E=e) = \sum_{x \in \mathcal{X}} \sum_{y \in \mathcal{Y}} p_{XY}(x,y|E=e) \log \left( \frac{p_{XY}(x, y|E=e)}{p_{X}(x|E=e) p_{Y}(y|E=e)} \right)
\end{equation}
This is not done if the parent variable, in the $X \rightarrow Y$ tuple under consideration, is in the evidence set; this is because observing a variable in a BN conceptually corresponds to disconnecting it from its children.

@@

>>>>

\todo{la discussione con l'esperto ci da' belief revision dei dati}
\todo{counterfactual explanations}
\todo{magari togliere theory e mettere tutto assieme}

\subsection{Entropy-Based Selection} \label{subsec:entropy-based-selection}
\todo{utilizzo entropia}

esempio fittizio che esemplifica il fatto che usare la probabilità è biased verso variabili a cardinalità più bassa
cardinalità diversa, grandezze non confrontabili
quanto ci si discosta dalla uniforme?  information gain, l'uniforme ha sempre entropia 1, con  cardinalita più grande a stessa probabilità abbiamo più informazione guadagnata

\subsection{Algorithms} \label{subsec:algorithms-novel}
An important part of my work was developing the algorithms needed to adapt the ideas presented in the paper \enquote{Explaining the Most Probable Explanation} by \cite{Butz2018} and \enquote{A Progressive Explanation of Inference in \enquote{Hybrid} Bayesian} by \cite{Kyrimi2016}.
From the former, the construction of the probability tree through a constructive dialogue with the domain expert, the building of counterfactual explanation branches, the automatic generation of the most probable probability tree from initial evidence.
From the latter, the generation of an \enquote{Inverse explanation}.
Finally, a simple procedure to output a natural language explanation was developed.

\subsubsection{Dialogues} \label{subsubsec:dialogues}
The so-called \enquote{Pseudo-MPE} algorithm is inherently wrapped up with the concept of \textit{dialogue} and is central to the explanatory powers of the system being developed in this thesis.
The algorithm was developed as a way of implementing the \enquote{MPE branch} of the \enquote{Argumentative Probability Tree} hypothesised by \cite{Butz2018}.
It was termed \enquote{Pseudo-MPE} because there are no guarantees of it returning the MPE solution (see Subsec. \ref{subsec:bnupdating}), as noted by \cite{koller2007introduction} in their definition of the MAP problem.

@@

\subsubsection{\enquote{Pseudo-MPE} from Initial Evidence} \label{subsubsec:pseudo-mpe}
In order to compare the Pseudo-MPE output with the true MPE solution I implemented a simple algorithm that, starting from an initial set of evidence, generates the relative pseudo-MPE solution.
The random initial evidence set is constructed by randomly choosing a number in the interval $k = [ 1, |V| ]$, with $V$ the set of vertices in the BN, and then randomly selecting $k$ of the random variables in $V$ to yield the set of variables $E$.
The value of each variable is randomly chosen among the set of its values; as all variables are categorical, this can easily be done.

An optional threshold can be set in order to discard $(state,value)$ pairs whose probability is deemed too low.

@@

The pseudocode is shown in Alg. \ref{alg:pseudo-mpe-initial-evidence}.

\begin{algorithm}[htp!]
	\caption{Pseudo-MPE from inital evidence algorithm}
	\label{alg:pseudo-mpe-initial-evidence}
	\begin{algorithmic}[1]
		\State $evidence$ defined by user
		\State $threshold$ defined by user
		\State $MPE\_polytree = $ MPE Polytree rooted in $evidence$
		\State $last\_node = evidence$
		\State $alternative\_evidence = evidence$ 
		\While{True} 
			\State $mpe\_states$ = \texttt{next\_most\_probable\_states($alternative\_evidence$)}
			\If{$mpe\_states$ is empty}
				\State return
			\Else
				\State $next\_state$ = head of $mpe\_states$ 
				\If{probability of $next\_state \leq threshold$}
					\State return constructed $MPE\_polytree$
				\EndIf
				\State create $next\_state$ node and make it son of $last\_node$
				\State update $alt\_node$ node to be $next\_state$ node
				\State $alternative\_evidence = alternative\_evidence \cup next\_state$
			\EndIf
		\EndWhile
	\end{algorithmic}
\end{algorithm}

\subsubsection{MPE Algorithms Comparison}
In order to compare the quality of the solution found using the simple Pseudo-MPE heuristic, I set up an experiment to compare it to the exact MPE solutions calculated with DAOOPT (see Subsec. \ref{subsec:libraries}).
The user is able to select the number of iterations over which to average the results and at each iteration a Pseudo-MPE solution is generated using Alg. \ref{alg:pseudo-mpe-random-evidence}.
The same initial evidence is fed to DAOOPT using the \texttt{daoopt\_solver} interfacing procedure shown in Subsec. \ref{subsec:algorithms} and to pgmpy's \texttt{map\_query} function.

At each iteration corresponds a random evidence; the solutions given by the Pseudo-MPE, pgmpy and by DAOOPT are scored against each other using Hamming (Subsec. \ref{subsec:hamming-distance}) and Jaccard (Subsec. \ref{subsec:jaccard-distance}) distances on the vectors representing the returned assignments.
The sequence of distances is then averaged over the iterations to return the two averaged distances between the three solutions to the MPE problem.

The implementation of the algorithm had to deal with the fact that DAOOPT failed on certain configurations of input evidence.
The reason for this is unknown so my workaround was to discard any iterations where DAOOPT had failed and retry them with another random set of evidences.
The pseudocode for the algorithm is shown in Alg. \ref{alg:mpe-comparison}.

\begin{algorithm}[htp!]
	\caption{MPE comparison algorithm}
	\label{alg:mpe-comparison}
	\begin{algorithmic}[1]
		\State user defined number of $iterations$
		\State initialise empty $results$ list
		\While{$i < iterations$}
			\State generate random number $k \in [ 1, |V| ]$
			\State $S = $ choose $k$ variables from $V$
			\For{$s \in S$}
				\State choose random $v$ in the possible values of $e$ 
				\State $evidence = evidence \cup (s,v)$
			\EndFor
			\State $pseudo\_mpe = $ \texttt{generate\_pseudo\_mpe($evidence$)}
			\State $model\_uai=$\texttt{export\_model\_to\_uai()}
			\State $evidence\_uai=$\texttt{export\_evidence\_to\_uai()}
			\State $daoopt\_mpe = $ \texttt{daoopt\_solver($model\_uai$, $evidence\_uai$)}
			\If{$daoopt\_mpe$ == -1} \Comment{DAOOPT has failed to calculate a solution}
				\State cancel $pseudo\_mpe$
				\State continue
			\EndIf
			\State compute $d_H(pseudo\_mpe,daoopt\_mpe)$, $d_J(pseudo\_mpe,daoopt\_mpe)$ and store in $iter\_result$
			\State append $iter\_result$ to $results$
			\State $i+=1$
		\EndWhile
		\State compute average in $results$ for $d_H$ and $d_J$
	\end{algorithmic}
\end{algorithm}

@@

\subsubsection{Pairwise Correlations}
An interesting addition, in terms of both explainability and theory, is an algorithm to measure and graphically display the interrelatedness between pairs of variables.
This is achieved by calculating the \textit{conditional mutual information} between each pair of parent $\rightarrow$ child variables.
The definition of mutual information presented in Subsec. \ref{subsec:mutualinformation} is extended to account for the current state of the model i.e. the set of observed variables.
What we are calculating is:
\begin{equation}
	I(X,Y|E=e) = \sum_{x \in \mathcal{X}} \sum_{y \in \mathcal{Y}} p_{XY}(x,y|E=e) \log \left( \frac{p_{XY}(x, y|E=e)}{p_{X}(x|E=e) p_{Y}(y|E=e)} \right)
\end{equation}
This is not done if the parent or child variable, in the $X \rightarrow Y$ tuple under consideration, is in the evidence set; this is because observing a variable in a BN conceptually corresponds to disconnecting it from its children.

@@

\subsection{Interfacing with the User}
Natural language was believed to be the most spontaneous way to make the expert user aware of the knowledge in the data set and for her to interact with the system.
Thus the main interaction mode with the implemented system is in using natural language, both as input and as output.
Natural language denominators come \enquote{for free} when dealing with the node names and their values, regardless of their internal representation in the implemented model, as these are simply the column names and values found in the original input data set.

A bit more care needs to be taken when quantifying probabilities to the user.
The format chosen should fit in as seamlessly as possible with the rest of the phrases generated and be able to conform to the user's preconceptions regarding probabilities.
It should be both \textit{informative} and \textit{precise}.

The chosen coding was taken from \cite{Butz2018} and is shown in Tab. \ref{tab:naturallanguageprobabilities}.
The mapping is used every time there is a need to translate a probability into natural language to be output to the user; for example when proposing tuples during a Dialogue (see Subsec. \ref{subsubsec:pseudo-mpe}) as example of which can be seen in Fig. \ref{fig:nl-dialogue}.

In Fig. \ref{fig:nl-independencies}, \ref{fig:nl-conditional} and \ref{fig:nl-mpe} we can see how the interface to the system's functions leans heavily on natural language elements.
The design ideal was for the system to be as accessible and the least intimidating possibile, in order to let the user extract the maximum amount of information in an easy way.
Not only the terms but also the phrasing were deemed to be important: every interaction is broken down into what are though to be manageable steps.
For example, a user isn't asked to provide all evidence at once but is asked for one at a time, when necessary.
The idea is to reduce cognitive overload on the expert's part so that he may be able to better concentrate on the task at hand, not on the tool he is using to achieve it.

Another important aspect is to translate the information into a language that would make semantic sense to the user; that is it should be part of the user's \textit{ontology}, the set of concepts part of his World.
This can be seen in Fig. \ref{fig:nl-independencies} where the notion of d-separation in the underlying BN's DAG is remapped to a higher-level concept that is part of the user's ontology, like that of variables affecting each other's values or not.
The underlying theme of the work carried out in this thesis is to translate \textit{information} present in a data set into \textit{knowledge} for a medical expert user.
This is done in various ways, for example through the use of dialogue, to bring the data to a form that is manageable and comprehensible.

A final element is that of a coherent use of colours, even in the experimental command-line interface that has currently been developed.
In the generated phrases the source variables are always highlighted in magenta, the evidences in green and the results of the queries in cyan.
It is hoped that this guide the user's eye towards the important elements of what he had set out to achieve, thus anchoring his experience and letting the marginal elements fade into background over time.

The grammars for the generation of the natural language are now presented, defined in Extended Backus-Naur form (EBNF).

\begin{algorithm}[htp!]
	\caption{Grammar generating dialogue output}
	\label{alg:nl-dialogue}
	\begin{algorithmic}[1]
		\State Next\_state := 'var 1' | $\ldots$ |  'var n'
		\State Value := 'val 1' | $\ldots$ |  'val k'
		\State Probability := 'highly unlikely' | 'very unlikely' | 'unlikely' | 'not plausibly' | 'plausibly' | 'possibly' | 'likely' | 'very likely' | 'highly likely' | 'certain'
		\State Output := 'The next most probable state is ' Probability Next\_state 'Enter to accept or n to refuse: ' (Still\_to\_explain ',')* | 'Is state ' Next\_state ' with value ' Value ' more correct?' 
		\State Still\_to\_explain := 'var 1' | $\ldots$ |  'var n'
	\end{algorithmic}
\end{algorithm}

\begin{algorithm}[htp!]
	\caption{Grammar generating independencies query output}
	\label{alg:nl-independencies}
	\begin{algorithmic}[1]
		\State Source := \#user input\#
		\State Evidence := 'var 1' | $\ldots$ |  'var n'
		\State Separated := 'var 1' | $\ldots$ |  'var n'
		\State Output := 'Given source variable' Source 'and given evidence: '  (Evidence ',')* 'the following variables have no effect: ' (Evidence ',')*
	\end{algorithmic}
\end{algorithm}

\begin{algorithm}[htp!]
	\caption{Grammar generating conditional probability query output}
	\label{alg:nl-conditional}
	\begin{algorithmic}[1]
		\State Source := \#user input\#
		\State Evidence := 'var 1' | $\ldots$ |  'var n'
		\State Output := 'Given target variable' Source 'and observed evidence: ' (Evidence 'with value: ' Value)* 'then the predicted values for ' Source 'are: ' (Evidence 'with value: ' Value)*
	\end{algorithmic}
\end{algorithm}

\begin{algorithm}[htp!]
	\caption{Grammar generating MPE query output}
	\label{alg:nl-mpe}
	\begin{algorithmic}[1]
		\State Source := \#user input\#
		\State Evidence := 'var 1' | $\ldots$ |  'var n'
		\State Value := 'val 1' | $\ldots$ |  'val k'
		\State Output := 'Given observed evidence:' (Evidence 'with value: ' Value)* 'the most probable configuration of the other variables is: ' (Evidence 'with value: ' Value)*
	\end{algorithmic}
\end{algorithm}

One of the most important parts of this work is the validation of this interface with real medical expert users, the results of which are presented in chap. \ref{chap:results}.

\begin{table*}[htbp]
\centering
\caption{Probability quantifiers in natural language}
\begin{tabularx}{\textwidth/2}{ X X }
\toprule 
Probability range & Natural language quantifier \\
\midrule 
(0, 0.2) &  "highly unlikely" \\
(0.2, 0.3) & "very unlikely" \\
(0.3, 0.4) & "unlikely" \\
(0.4, 0.5) & "not plausibly" \\
(0.5, 0.6) & "plausibly" \\
(0.6, 0.7) & "possibly" \\
(0.7, 0.8) & "likely" \\
(0.8, 0.9) & "very likely" \\
(0.9, 1) &  "highly likely" \\
(1) &  "certain" \\
\bottomrule
\end{tabularx}
\label{tab:naturallanguageprobabilities}
\end{table*}

\begin{figure}[htbp]
\centerline{\includegraphics[width=\columnwidth]{methodology/images/nl-dialogue}}
\caption{Interface while executing a Dialogue}
\label{fig:nl-dialogue}
\end{figure}

\begin{figure}[htbp]
\centerline{\includegraphics[width=\columnwidth]{methodology/images/nl-independencies-query}}
\caption{Interface while executing a query on the d-separations}
\label{fig:nl-independencies}
\end{figure}

\begin{figure}[htbp]
\centerline{\includegraphics[width=\columnwidth]{methodology/images/nl-conditional-query}}
\caption{Interface while executing a conditional probability query}
\label{fig:nl-conditional}
\end{figure}

\begin{figure}[htbp]
\centerline{\includegraphics[width=\columnwidth]{methodology/images/nl-mpe-query}}
\caption{Interface while executing an MPE query}
\label{fig:nl-mpe}
\end{figure}

### 2019-08-06 methodology/validation.tex
Literature Review

\section{Validation}
\todo[inline]{discutere con Vittoria come validare al meglio}
Having direct access to expert pathologists has not only helped in guiding research into the theoretical explainability properties of the system but also enabled their validation.
There are two main validation streams to be addressed: from the clinical point of view and from the explainability one, with the results of the latter depending on those of the former.

\subsection{Clinical validation} 
A validation of the methods carried out in this thesis in their adherence to established clinical literature, is of paramount importance.
A failure on the Bayesian Network's part in capturing the true relationships between the variables would hamper it in being able to give any meaningful representation of them.
For the expert to even start to trust the system or to be able to make sense of its outputs, it is vital that there be as little cognitive dissonance between the experts' basic beliefs and expectations and those that he sees represented in the system.

For this reason, the initial validation phase with the Istituto Cantonale di Patologia concentrated on the clinical validation aspect.
The methodology chosen to clinically validate the system was for a representative of the ICT to formulate a series of natural language queries; each one of these was annotated with the specific queried variable in the network and its value together with the known values of other variables.
The ICT's delegate included the expected reply to the queries together with its probability, based on the latest medical literature and their personal expertise.
So, each query asked for the probability truth value of the following:
\begin{align}
	(var_1 = a \wedge \ldots \wedge var_n = z ) \Rightarrow var_k = k \quad var_1 \ldots var_n \in V, var_k \in V \smallsetminus \{ var_1 \ldots var_n \}
\end{align}
This would translate into natural language into:
\enquote{If $var_1$ is $a$ and $\ldots$ and $var_n$ is $z$, how probable is it that $var_k$ $k$?}.
The complete series of eight questions, together with the expected values, is shown in Tab. \ref{tab:clinicalvalidationquestions}.
It is quite evident that all these validation questions are instances of the Bayesian Network Updating problem, in particular they are Conditional Probability Queries (as defined in \ref{subsec:bnupdating}).

@@

\subsection{Explainability validation} \label{subsec:explainability-validation}
Having validated the system in terms of its adherence to clinical literature, it could then also meaningfully be validated from an explainability point of view.
The main question to be addressed is its capacity to relate to the expert user.
Is the system able to engender the user's trust?
In doing so, is she able to extract more knowledge from existing data when using the system than not?
Especially in cases where there may be a dearth of data, can the expert maximise the benefit from the available information?
Does the user subjectively feel that the system may positively impact her work?
These are all hard questions to answer, as there is a very high degree of subjectivity involved.
Thus to attempt to answer them, the chosen methods were borrowed from the social sciences.

In an earlier stage, the experts were introduced to the system in prototype form and instructed on the possible use cases it offered.
Then, they were asked to keep track of times they felt they could have used the system, had it been available, during their day-to-day work and to report back to me with the query they would have submitted to the system.
If the query were in the realm of possibilities offered by my program, I would submit the outputs back to the users together with the following questionnaire:
\todo[inline]{definire questionario da mandare con le domande}
This process would both give feedback on the performance of the system and also inform its design.

In a later phase, having finalised the system's design, I provided it to the experts at the ICT for their use in their daily work.
After \textbf{xx time} I followed up with them with an interview based on the following format:
\todo[inline]{definire formato e metriche intervista}
This enabled me to quantify the performance of the system, as perceived by its users in a real setting, over an extended period of time.

>>>>

\section{Validation} \label{sec:validation}
\todo[inline]{discutere con Vittoria come validare al meglio}
Having direct access to expert pathologists has not only helped in guiding research into the theoretical explainability properties of the system but also enabled their validation.
There are two main validation streams to be addressed: from the clinical point of view and from the explainability one, with the results of the latter depending on those of the former.

\subsection{Clinical validation} \label{subsec:clinical-validation-methodology}
A validation of the methods carried out in this thesis in their adherence to established clinical literature, is of paramount importance.
A failure on the Bayesian Network's part in capturing the true relationships between the variables would hamper it in being able to give any meaningful representation of them.
For the expert to even start to trust the system or to be able to make sense of its outputs, it is vital that there be as little cognitive dissonance between the experts' basic beliefs and expectations and those that he sees represented in the system.

For this reason, the initial validation phase with the ICP concentrated on the clinical aspect.
The methodology chosen to clinically validate the system was represented by a series of queries, formulated in natural language by the ICP; each one of these questions was annotated with the specific queried variable in the network and its value together with the known values of other variables.
The experts included the expected reply to the queries together with its probability, based on the latest medical literature and their personal, knowledge-based expertise.
So, each query asked for the probability truth value of the following:
\begin{align}
	(var_1 = a \wedge \ldots \wedge var_n = z ) \Rightarrow var_k = k \quad var_1 \ldots var_n \in V, var_k \in V \smallsetminus \{ var_1 \ldots var_n \}
\end{align}
This would translate into natural language into:
\enquote{If $var_1$ is $a$ and $\ldots$ and $var_n$ is $z$, how probable is it that $var_k$ $k$?}.
The complete series of eight questions, together with the expected values, is shown in Tab. \ref{tab:clinicalvalidationquestions}.
It is quite evident that all these validation questions are instances of the Bayesian Network Updating problem, in particular they are Conditional Probability Queries (as defined in \ref{subsec:bnupdating}).

@@

\subsection{Explainability validation} \label{subsec:explainability-validation}
In general, there is strong resistance to novelty in the field of Medicine, both because of ethical reasons and because of the necessity for clinicians to be conservative in attending to best practices.
Any tool that is too onerous in terms of time and cognitive load is liable to remain unutilised.
In this field, a tool must only be the means by which a question is answered, not become a question itself; the system developed in this thesis aims to conform to this objective.
The need for a comprehensible and efficient tool is especially present because the goal of a pathologist is to arrive at a diagnosis, containing the elements useful to define prognosis and therapeutical approach, in the briefest time possible.
The main reasons are ethical, as for a patient waiting for a report is extenuating, and clinical, because tempestivity in diagnosis is the first factor at the base of life expectancy.
Obviously, the highest accuracy possible is always strived for.
The clinical field and that of biomedicine are forced to embrace uncertainty, as this is an integral part of their practice.
Consequently, any tool able to support in comprehension and decision-making is automatically useful, once it has been clinically validated; in other words, even though a specific system may not be decisive or applicable to all reviewed cases, it will nonetheless be taken into account.

Thus, after having validated the system in terms of its adherence to clinical literature, it could then also meaningfully be validated from an explainability point of view.
The main question to be addressed is its capacity to relate to the expert user.
Is the system able to engender the user's trust?
In doing so, is she able to extract more knowledge from existing data when using the system than not?
Especially in cases where there may be a dearth of data, can the expert maximise the benefit from the available information?
Does the user subjectively feel that the system may positively impact her work?
These are all hard questions to answer, as there is a very high degree of subjectivity involved.
Thus to attempt to answer them, the chosen methods were borrowed from the social sciences.

In an earlier stage, the experts were introduced to the system in prototype form and instructed on the use cases it offered.
This process would give feedback on the functionalities of the system and help in shaping its design.

The finalised system was, in a later phase, provided to the experts at the ICP for use in their daily work.
To quantify the performance of the system, as perceived by its users in a real setting over an extended period of time, a follow-up was done after \textbf{xx time} by way of the following questionnaire, designed to test the gaps identified in Chap. \ref{chap:literature-review} in mind:
\todo{inserire questionario}
